---
title: "Data Manipulation"
author: "Kallan Roan"
date: "2025-09-18"
output: github_document
---

```{r}
library(tidyverse)
```

Let's import the dataset
```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))
litters_df = 
  janitor::clean_names(litters_df)

pups_df = 
  read_csv(
    "data/FAS_pups.csv", 
           skip = 3, 
           na = c("NA", ".", ""))

pups_df = 
  janitor::clean_names(pups_df)
```

## 'select'

'select' selects variables form my dataframe.

select specific of variables
```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range
```{r}
select(litters_df, group:gd_of_birth)
```

select by removal

remove the group variable
```{r}
select(litters_df, -group)
```
remove a range of variables

```{r}
select(litters_df, -(group:gd_of_birth))
```

select by prefix

```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting

```{r}
select(litters_df, starts_with("gd"), group)
```
give me litter_number, and then give me everything else in the datset
```{r}
select(litters_df, litter_number, everything())
```

does the same thing as select

```{r}
relocate(litters_df, litter_number)
```

rename by selecting
```{r}
select(litters_df, GROUP = group, everything())
rename(litters_df, GROUP = group)
```

What if I only select one variable

```{r}
select(litters_df, group)

```
Rarely use pull to pull something out of a dataframe
get a vector

```{r}
pull(litters_df, group)
```


Learning Assessment: In the pups data, select the columns containing litter number, sex, and PD ears.
```{r}
select(pups_df, litter_number, sex, pd_ears )
```




