---
title: "Data Manipulation"
author: "Kallan Roan"
date: "2025-09-18"
output: github_document
---

```{r}
library(tidyverse)
```

Let's import the dataset
```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))
litters_df = 
  janitor::clean_names(litters_df)

pups_df = 
  read_csv(
    "data/FAS_pups.csv", 
           skip = 3, 
           na = c("NA", ".", ""))

pups_df = 
  janitor::clean_names(pups_df)
```

## 'select'

'select' selects variables form my dataframe.

select specific of variables
```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range
```{r}
select(litters_df, group:gd_of_birth)
```

select by removal

remove the group variable
```{r}
select(litters_df, -group)
```
remove a range of variables

```{r}
select(litters_df, -(group:gd_of_birth))
```

select by prefix

```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting

```{r}
select(litters_df, starts_with("gd"), group)
```
give me litter_number, and then give me everything else in the datset
```{r}
select(litters_df, litter_number, everything())
```

does the same thing as select

```{r}
relocate(litters_df, litter_number)
```

rename by selecting
```{r}
select(litters_df, GROUP = group, everything())
rename(litters_df, GROUP = group)
```

What if I only select one variable

```{r}
select(litters_df, group)

```
Rarely use pull to pull something out of a dataframe
get a vector

```{r}
pull(litters_df, group)
```


Learning Assessment: In the pups data, select the columns containing litter number, sex, and PD ears.
```{r}
select(pups_df, litter_number, sex, pd_ears )
```


## `filter`

use `filter()` to filter out rows

keep rows that have gd_of_birth of 20 and pups born alive = 6
```{r}
filter(litters_df, gd_of_birth == 20, pups_born_alive == 6)
```

look for more than two pups alive, more than or equal to, less than or equal to...

```{r}
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive >= 5)
filter(litters_df, pups_born_alive <= 5)

```

look for non-matches

```{r}
filter(litters_df, pups_born_alive != 6)
```


match to a set

look for groups that are in Con7 and Con8

```{r}
filter(litters_df, group %in% c("Con7", "Con8"))
```

filter out missing rows with `drop_na()`

```{r}
drop_na(litters_df)
```

only filter out missing rows in one variable only (gd0_weight)

```{r}
drop_na(litters_df, gd0_weight)
```



Learning Assessment: In the pups data:
Filter to include only pups with sex 1
Filter to include only pups with PD walk less than 11 and sex 2

```{r}
filter(pups_df, sex == 1)
filter(pups_df, pd_walk < 11, sex == 2)
```


## `mutate`

use `mutate()` to create or modify variables


create wt_gain variable
make all results under group variable lowercase

```{r}
mutate(
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
  )
```

## `arrange`

`arrange()` arranges things

organize based on variable pups_born_alive

```{r}
arrange(litters_df, pups_born_alive)
```

Arrange by group variable first, and then by pups_born_alive

```{r}
arrange(litters_df, group, pups_born_alive)
```

Put pups_born_alive in descending order
```{r}
arrange(litters_df, desc(pups_born_alive))
```


## PIPES

Let's see a bad way to do things. Don't do this

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df = 
  janitor::clean_names(litters_df)

litters_df =
  select(litters_df, group, litter_number, starts_with("gd"))

litters_df = 
  drop_na(litters_df)

litters_df = 
  mutate(
    litters_df,
    wt_gain = gd18_weight - gd0_weight
  )
```

Use pipes to avoid all of this

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  select(group, litter_number, starts_with("gd")) |> 
  drop_na() |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight
  )
```

Learning Assessment: Write a chain of commands that:
loads the pups data
cleans the variable names
filters the data to include only pups with sex 1
removes the PD ears variable
creates a variable that indicates whether PD pivot is 7 or more days

```{r}
pups_df = 
  read_csv(
    "data/FAS_pups.csv", 
    skip = 3, 
    na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(
    pd_pivot_ge7 = pd_pivot >= 7
  )
```

